---
title: "iMKT using PopFly or PopHuman data"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    highlight: tango
    code_folding: show
bibliography: references.bib
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{iMKT Pipeline}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	fig.height = 7,
	fig.width = 7,
	collapse = TRUE,
	comment = "#>"
)
```

The improvement of NGS technologies that lead to the sequencing of hundreds or even thousands of individuals from the same species provides us with the necessary data on which to empirically test evolutionary theoretical hypotheses. The availability of such datasets increases the request of developing high performance specific tools able to deal and analyze this information.

The PopFly [@popfly_2017] and PopHuman [@pophuman_2018] genome browsers are the two central repositories of population genomics estimates for *Drosophila melanogaster* and *Homo sapiens* model species, respectively. Thus, we decided to provide functionalities to directly link the iMKT software to both databases, which allows retrieving and analyzing population genomics information in a single step.

Therefore, iMKT package includes some functions which allow an easy retrieval and analysis of population genetics information stored in these genome browsers. Specifically, the functions permit the download of population genetics parameters computed for every gene annotation in several populations for both model species.

This vignette is divided in two main sections:

- Loading the raw data
- Performing iMKT analyses

Although the examples presented here focus only on PopFly data and analyses, the same process could be performed using human data and functions (**PopHumanData**, **loadPopHuman()**, **PopHumanAnalysis()**), following the same steps described in this document. Recombination rate values included in human data were retrieved from @bherer_2017 and correspond to the sex average estimates, while for *D. melanogaster* comes from @comeron_2012.

&nbsp;

## Loading the raw data

The first step is to load the information into your working environment. This would allow a manual examination of the data before starting any analysis. However, to speed analyses, this step can be skipped and the main data object is loaded into the workspace the first time that the corresponding analysis function is called. To download PopFly data use the **loadPopFly()** function, without any argument. 

Keep in mind that you are downloading full-genome information of several populations (13,753 gene annotations for 16 populations in *D.melanogaster* and 20,643 gene annotations for 26 populations in *H. sapiens*), and this process could take a while.

Once the function finishes,the **PopFlyData** object is loaded into the workspace. This object can be manually examined or used when performing iMKT analyses.
  
&nbsp;

```{r popfly data, echo=TRUE, fig.width=7, messages=F, warning=FALSE}
## Load the iMKT library
library(iMKT)

## Load PopFlyData
loadPopFly()

## Check data object
ls()
names(PopFlyData)
```

&nbsp;
  
Each row of the **PopFlyData** dataframe contains information regarding one gene annotation in one single population. Metrics for each gene contain information about the number of segregating, divergent and analyzed positions, and the Derived Allele Frequency distribution (DAF) for neutral (4fold) and putatively selected (0fold) sites; together with some neutrality tests statistics (Standard MKT, Direction of Selection, Ka/Ks) and gene-associated recombination rate estimates (retrieved from @comeron_2012).

Once the data is loaded into the workspace, diverse iMKT analyses can be performed using the function **PopFlyAnalysis()**.

&nbsp;
&nbsp;

## Performing iMKT Analyses
  
The **PopFlyAnalysis()** function allows performing any MK test using a subset of PopFly data defined by custom genes and populations lists. It uses the previously loaded dataframe (PopFlyData). In addition to the **genes** and **populations** lists, the function also has the following parameters:

  - **recomb**: group genes according to recombination values (TRUE/FALSE)
  - **bins**: number of recombination bins to compute (mandatory if `recomb=TRUE`)
  - **test**: which test to perform. Options include: `standardMKT` (default), `imputedMKT`, `eMKT`, `FWW`, `aMKT`
  - **xlow**: lower limit for asymptotic alpha fit (default=0)
  - **xhigh**: higher limit for asymptotic alpha fit (default=1)
  - **plot**: report plot (optional, default=FALSE)

Custom genes must be listed using FlyBase IDs (FBgn...), and the available populations from PopFly are: AM, AUS, CHB, EA, EF, EG, ENA, EQA, FR, RAL, SA, SD, SP, USI, USW, ZI.

Hence, using the parameters `recomb` and `bins`, the function allows deciding whether to analyze genes grouped by recombination bins or not.

&nbsp;

### Example 1. imputedMKT analysis without recombination
  
In this first example, the analysis is focused in two genes and two populations, without considering gene's recombination context, and using the `imputedMKT` methodology. 

The function groups polymorphism and divergence values of the custom genes, creating a new "concatenated gene" for each population of interest and performs the test defined. Then, it returns a list of lists with the default test output (`imputedMKT` in this case) for each population (RAL and ZI).

The output contains the standard MKT table, the $\alpha$ estimate (and its
associated p-value) for the chosen cutoff, and the negative selection
fractions $d$/$f$/$b$ (see the iMKT Pipeline vignette for details on each
of these). Note: `eMKT()` can be used the same way (`test="eMKT"`) and
reports a genuinely different $\alpha$ estimate (see the
[iMKT Pipeline](iMKTPipeline.html) vignette for the distinction between the two).

&nbsp;

```{r PopFly data retrieve automated no recomb, fig.height = 10, echo=TRUE}
PopFlyAnalysis(genes=c("FBgn0000055","FBgn0003016"), pops=c("RAL","ZI"), recomb=F, test="imputedMKT", plot=TRUE)
```
  
&nbsp;
&nbsp;
  
### Example 2. aMKT analysis using recombination bins
 
In this second example, genes from RAL population are grouped in three recombination bins, using recombination values from @comeron_2012. The test used is `aMKT`, with `xlow` and `xhigh` values set to 0 and 0.9, respectively. We are analyzing the complete set of genes located in the chromosome 2R ($n=2,822$)

In this case, the function creates three different "concatenated" genes (941, 940 and 941 genes, respectively), grouping again polymorphism and divergence values. These aggrupations are made according to the gene associated recombination rate estimates. The function returns a list of lists with the default test output (`aMKT` in this case) for each population (RAL) and recombination bin (1, 2 and 3).

The output of the function is a list that contains:

- **Asymptotic MK table**: table including information about the model type (exponential) along with the fitted function values ($a$, $b$, $c$), the $\alpha$ asymptotic estimate with its corresponding lower and higher confidence interval values, and the $\alpha$ original estimate (using the standard MKT methodology and the polymorphic sites within the `xlow` and `xhigh` cutoffs)
- **Fractions of sites**: negative selection fractions ($d$: strongly deleterious, $f$: neutral and $b$: weakly deleterious)
- **Graphs**: 3 plots showing **(A)** the distribution of allele frequencies (DAF) for neutral and selected sites, **(B)** the adaptation values ($\alpha$) for each DAF category along with the function fit, the $\alpha$ asymptotic and $\alpha$ original estimates and the limits used for function fitting and adaptation values calculation, and **(C)** the negative selection fractions

&nbsp;

```{r PopFly data retrieve automated, echo=TRUE, fig.height = 10}
geneList <- as.vector(unique(PopFlyData[PopFlyData$Chr=="2R",]$Name))
PopFlyAnalysis(genes=geneList , pops="RAL", recomb=T, bins=3, test="aMKT", xlow=0, xhigh=0.9, plot=TRUE)
```
  
&nbsp;

## References